---
title: "Displaying Distributions of Continuous Data With 5 Different Charts"
description: |
  Histogram, Density, Violin, Boxplot, Ridgeline.
author:
  - name: A. Uraz Akgül
date: 2022-08-28
output:
  distill::distill_article:
    self_contained: false
categories:
  - Data Visualization
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Borsa Istanbul calculates variety of indices to follow the movements of markets, to be an underlying asset for the financial products, and to be used as a benchmark for collective investment instruments.* Click [here](https://borsaistanbul.com/en/sayfa/2818/bist-indices) for further information.

You can access the data by downloading *post21.xlsx* file on [my GitHub account](https://github.com/rpydaneogrendim/rblog/tree/main/data). The 12 BIST Indices were chosen by me. You can find all of BIST indices [here](https://www.kap.org.tr/en/Endeksler). Reuters data is used.

```{r}

library(tidyverse)

df <- readxl::read_excel("data.xlsx") %>% 
  select(-DATE) %>% 
  mutate_all(
    ., function(x) (lead(x)/x-1)*100
  ) %>% 
  na.omit()

```

```{r eval=FALSE}

head(df)

```

```{r echo=FALSE}

df %>% 
  mutate_all(., function(x) round(x, digits = 2)) %>% 
  head(.) %>% 
  kableExtra::kbl() %>% 
  kableExtra::kable_styling()

```

```{r}

df <- df %>% 
  mutate(t = seq(1,nrow(.),1), .before = 1) %>% 
  pivot_longer(!t, names_to = "vars", values_to = "vals")

```

**Histogram and Density**

If there are two variables:

```{r fig.width=10, fig.height=6}

df %>% 
  filter(vars %in% c("XU100","XBANK")) %>% 
  ggplot(aes(x = vals, fill = vars)) +
  geom_histogram(position = "identity", alpha = .5) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        legend.title = element_blank(),
        legend.position = "top") +
  scale_fill_manual(values = c("red","blue"))

```

```{r fig.width=10, fig.height=6}

df %>% 
  filter(vars %in% c("XU100","XBANK")) %>% 
  ggplot(aes(x = vals, fill = vars)) +
  geom_density(alpha = .5) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        legend.title = element_blank(),
        legend.position = "top") +
  scale_fill_manual(values = c("red","blue"))

```

```{r fig.width=10, fig.height=6}

df %>% 
  filter(vars %in% c("XU100","XBANK")) %>% 
  pivot_wider(names_from = "vars", values_from = "vals") %>% 
  ggplot(aes(x = vals)) +
  geom_histogram(aes(x = XU100, y = ..density..), fill = "red" ) +
  geom_label(aes(x = 4, y = 0.25, label = "XU100"), color = "red") +
  geom_histogram(aes(x = XBANK, y = -..density..), fill = "blue") +
  geom_label(aes(x = 4, y = -0.25, label = "XBANK"), color = "blue") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank())

```

```{r fig.width=10, fig.height=6, preview=TRUE}

df %>% 
  filter(vars %in% c("XU100","XBANK")) %>% 
  pivot_wider(names_from = "vars", values_from = "vals") %>% 
  ggplot(aes(x = vals)) +
  geom_density(aes(x = XU100, y = ..density..), fill = "red" ) +
  geom_label(aes(x=4, y=0.25, label = "XU100"), color="red") +
  geom_density(aes(x = XBANK, y = -..density..), fill = "blue") +
  geom_label(aes(x=4, y = -0.25, label = "XBANK"), color="blue") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank())

```

If there are more than two variables:

```{r fig.width=10, fig.height=6}

df %>% 
  filter(vars %in% c("XU100","XBANK","XBLSM")) %>% 
  ggplot(aes(x = vals, color = vars)) +
  geom_density(lwd = 1.5) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        legend.title = element_blank(),
        legend.position = "top")

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vals, fill = vars)) +
  geom_histogram(position = "identity") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none") +
  facet_wrap(~vars)

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vals, fill = vars)) +
  geom_density() +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none") +
  facet_wrap(~vars)

```

**Violin and Boxplot**

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vars, y = vals, fill = vars)) +
  geom_violin() +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none")

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vars, y = vals, fill = vars)) +
  geom_violin() +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none") +
  coord_flip()

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vars, y = vals, fill = vars)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none")

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vars, y = vals, fill = vars)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none") +
  coord_flip()

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vars, y = vals, fill = vars)) +
  geom_violin() +
  geom_boxplot(alpha = .5) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none")

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vars, y = vals, fill = vars)) +
  geom_violin() +
  geom_boxplot(alpha = .5) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none") +
  coord_flip()

```

**Ridgeline**

```{r}

library(ggridges)

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(x = vals, y = vars, fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none")

```

```{r fig.width=10, fig.height=6}

df %>% 
  ggplot(aes(vals, y = vars, fill = 0.5 - abs(0.5 - stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE) +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        legend.position = "none")

```