---
title: "How to Automate Repetitive and Time-Consuming Reports"
description: |
  Scheduling a report to run at regular intervals.
author:
  - name: Uraz Akgül
date: 2022-09-11
output:
  distill::distill_article:
    self_contained: false
categories:
  - Automation
  - Web
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the age of technology we live in, manual processes are extremely time-consuming. Fortunately, there are an increasing number of people who know how to use technology or desire to do so.

In this post, I'll show you how to run a script at regular intervals. Assume we want to get intraday cryptocurrencies data from Yahoo Finance every 5 minutes using web scraping. The top 100 cryptocurrencies in the table are what we are most interested in.

```{r}

library(rvest)
library(tidyverse)
library(DBI)
library(RSQLite)
library(taskscheduleR)

```

```{r}

url <- "https://finance.yahoo.com/cryptocurrencies/?count=100&offset=0"

df <- read_html(url) %>% 
  html_table() %>% 
  .[[1]]

```

The top ten cryptocurrencies in the table are listed below.

```{r echo=FALSE}

df %>% 
  head(.,10) %>% 
  kableExtra::kbl() %>% 
  kableExtra::kable_styling()

```

It's enough for us to select the Symbol and % Change columns.

```{r}

master <- df %>% 
  select(1,5) %>% 
  rename("Change"=2)

```

I'd like to draw your attention to the fact that the values in the Change column are not numeric format.

```{r}

str(master)

```

We need to remove the plus and percent signs from the column I mentioned and convert the values to numeric format.

```{r}

master <- master %>% 
  mutate(
    Change = as.numeric(gsub("[^0-9\\.]","",Change))
  )

```

There, it is done!

```{r}

str(master)

```

Before proceeding, I should point out that we will need to create a new column and get the dates as IDs as well.

```{r}

master <- master %>% 
  mutate(
    ID = Sys.time()
  )

```

And now we can move on to the data visualization step.

```{r}

gainers <- master %>% 
  arrange(desc(Change)) %>% 
  filter(Change > 0) %>% 
  slice(1:10)

if(nrow(gainers) > 0){
  g <- ggplot(gainers, aes(x = reorder(Symbol, Change), y = Change, fill = Change)) +
    geom_col() +
    coord_flip() +
    theme_minimal() +
    theme(
      plot.background = element_rect(fill = "#262626"),
      panel.background = element_blank(),
      panel.grid = element_line(color = "#4c4c4c"),
      axis.title = element_blank(),
      axis.text = element_text(color = "#ffffff", size = 15),
      legend.position = "none",
      plot.title = element_text(color = "#ffffff", size = 20, face = "bold"),
      plot.subtitle = element_text(color = "#ffffff", size = 13, face = "italic")
    ) +
    scale_fill_gradient(low = "red", high = "green") +
    labs(title = "Top 10 Gainers",
         subtitle = gainers$ID[1])
}

ggsave("g.png")

```

```{r echo=FALSE, fig.width=10, fig.height=7, preview=TRUE}

plot(g)

```

The following are the codes that we have written thus far.

```{r eval=FALSE}

library(rvest)
library(tidyverse)

url <- "https://finance.yahoo.com/cryptocurrencies/?count=100&offset=0"

df <- read_html(url) %>% 
  html_table() %>% 
  .[[1]]

master <- df %>% 
  select(1,5) %>% 
  rename("Change"=2) %>% 
  mutate(
    Change = as.numeric(gsub("[^0-9\\.]","",Change)),
    ID = Sys.time()
  )

gainers <- master %>% 
  arrange(desc(Change)) %>% 
  filter(Change > 0) %>% 
  slice(1:10)

if(nrow(gainers) > 0){
  g <- ggplot(gainers, aes(x = reorder(Symbol, Change), y = Change, fill = Change)) +
    geom_col() +
    coord_flip() +
    theme_minimal() +
    theme(
      plot.background = element_rect(fill = "#262626"),
      panel.background = element_blank(),
      panel.grid = element_line(color = "#4c4c4c"),
      axis.title = element_blank(),
      axis.text = element_text(color = "#ffffff", size = 15),
      legend.position = "none",
      plot.title = element_text(color = "#ffffff", size = 20, face = "bold"),
      plot.subtitle = element_text(color = "#ffffff", size = 13, face = "italic")
    ) +
    scale_fill_gradient(low = "red", high = "green") +
    labs(title = "Top 10 Gainers",
         subtitle = gainers$ID[1])
}

ggsave("g.png")

```

**Writing the data to the database.**

SQLite will be used in this study. My reasoning for choosing SQLite is that we will not require any software or server setup. It also has a simple structure. For SQLite, we will use the DBI and RSQLite packages, and the following command will be used to create the database (not temporary; permanent). Wherever the address is, the file path will save the database.

```{r}

# working directory
# getwd()

myDB <- dbConnect(SQLite(), "cryptoDB.sqlite") # "C:/.../cryptoDB.sqlite"

```

Let's write the data to the database named cryptoDB. When we do this on a regular basis, say every 5 minutes, we will be careful not to overwrite the table in the database with new data. The append parameter is set to TRUE for this purpose.

```{r}

dbWriteTable(myDB, "master", master, append = TRUE)

```

Let's get the data from the database using SQL.

```{r}

mastertbl <- dbGetQuery(myDB, "SELECT * FROM master")

```

```{r echo=FALSE}

mastertbl %>% 
  head(.,10) %>% 
  kableExtra::kbl() %>% 
  kableExtra::kable_styling()

```

When the processes are finished, use the code below to exit the database. It will be reconnected with the code that was written at the start (comment line below).

```{r}

dbDisconnect(myDB)

# myDB <- dbConnect(SQLite(), "cryptoDB.sqlite")

```

The following are the codes that we have written thus far.

```{r eval=FALSE}

library(DBI)
library(RSQLite)

myDB <- dbConnect(SQLite(), "cryptoDB.sqlite") # "C:/.../cryptoDB.sqlite"
dbWriteTable(myDB, "master", master, append = TRUE)
dbDisconnect(myDB)

```

**Schedule the task scheduler to run the script on a regular basis.**

The task scheduler can be configured via the PC or the taskscheduleR package.

```{r eval=FALSE}

taskscheduler_create(
  
  taskname = "Post27",
  rscript = "post27.R", # "C:/.../"
  schedule = "MINUTE",
  starttime = "12:00",
  modifier = 5
  
)

```

The process has been automated.

What can be done next?

* By converting the .R extension to the .Rmd extension, the report can be sent.

* E-mail can be used to send outputs with the .R or .Rmd extensions.