---
title: "Plotting Multiple Groups in ggplot2 Using facet_wrap() in Two Ways"
description: |
  How to split groups using facet_wrap() in ggplot2.
author:
  - name: A. Uraz Akgül
date: 2022-09-05
output:
  distill::distill_article:
    self_contained: false
categories:
  - Data Visualization
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this post, I'll show you how to split groups using facet_wrap() in ggplot2. I'd like to point out that I prefer the second method over the first because it is more informative.

You can access the data we'll be using by downloading the *post24.xlsx* file from [here](https://github.com/rpydaneogrendim/rblog/tree/main/data) or by following the steps below.

i. https://evds2.tcmb.gov.tr/index.php?/evds/serieMarket

ii. Interest Rates

iii. Weighted Average Interest Rates For Banks Loans (Flow Data, %)). Select the following items and click the Add button:

- Personal (TRY)(Flow Data, %)

- Vehicle (TRY)(Flow Data, %)

- Housing (TRY)(Flow Data, %)

- Commercial (TRY)(Flow Data, %)

iv. Date From: 01-01-2019, Date To: 26-08-2022. Click the Create Report icon under Report Settings.

v. You'll see the excel icon.

```{r}

library(tidyverse)

df <- readxl::read_excel("data.xlsx") %>% 
  pivot_longer(!date, names_to = "vars", values_to = "vals") %>% 
  mutate(date = lubridate::dmy(date))

```

**0. Before splitting groups with facet_wrap()**

```{r fig.width=15, fig.height=10}

ggplot(df, aes(x = date, y = vals, group = vars, color = vars)) +
  geom_line(size = 1) +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.title = element_blank(),
        legend.position = "top",
        legend.key.size = unit(1, 'cm'),
        legend.text = element_text(size = 20),
        axis.text = element_text(size = 20)) +
  labs(
    title = "Weighted Average Interest Rates For Banks Loans"
  )

```

**1. Splitting groups with facet_wrap()**

```{r fig.width=15, fig.height=10}

ggplot(df, aes(x = date, y = vals, group = vars, color = vars)) +
  geom_line(size = 1) +
  ggthemes::theme_fivethirtyeight() +
  theme(strip.text = element_text(size = 20),
        legend.position = "none",
        axis.text = element_text(size = 20)) +
  facet_wrap(~vars) +
  labs(
    title = "Weighted Average Interest Rates For Banks Loans"
  )

```

**2. Splitting groups with facet_wrap() using other variables**

When making comparisons between variables, I would prefer the second option most of the time.

```{r fig.width=15, fig.height=10, preview=TRUE}

ggplot(df, aes(x = date, y = vals)) +
  geom_line(data = df %>% rename(vars2 = vars), aes(group = vars2), color = "gray", size = 1) +
  geom_line(color = "dark blue", size = 2) +
  ggthemes::theme_fivethirtyeight() +
  theme(strip.text = element_text(size = 20),
        axis.text = element_text(size = 20)) +
  facet_wrap(~vars, ncol = 2) +
  labs(
    title = "Weighted Average Interest Rates For Banks Loans"
  )

```