---
title: "How Do Exchange Rates React to Interest Rate Changes?"
description: |
  Interest rate decisions by the CBRT and the reactions of the USDTRY.
author:
  - name: Uraz Akgül
date: 2022-10-09
output:
  distill::distill_article:
    self_contained: false
categories:
  - Finance
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

According to the information I obtained from [this](https://www.cbrates.com/) website, the central banks of the countries that raised and cut interest rates in 2022 are shown below.

```{r fig.width=10, fig.height=15}

library(rvest)
library(tidyverse)

url <- "https://www.cbrates.com/"

df <- read_html(url) %>% 
  html_table() %>% 
  .[[4]] %>% 
  slice(-c(1:2)) %>% 
  select(-c(1,4)) %>% 
  mutate(
    "Date" = lubridate::mdy(gsub("[\\(\\)]", "", regmatches(X3, gregexpr("\\(.*?\\)", X3)))),
    "Country" = str_squish(word(X3,1,sep = "\\|")),
    "Interest Rate" = as.numeric(word(X2,1,1)),
    "Bps" = str_squish(gsub("[\\(\\)]", "", regmatches(X2, gregexpr("\\(.*?\\)", X2)))),
    "Direction" = word(Bps,1,1),
    "Bps" = as.numeric(word(str_squish(gsub("[\\(\\)]", "", regmatches(X2, gregexpr("\\(.*?\\)", X2)))),2,2)),
    "Bps" = case_when(
      Direction == "+" ~ Bps*100,
      Direction == "-" ~ Bps*(-100)
    )
  ) %>% 
  select(3:7) %>% 
  na.omit() %>% 
  filter(Date >= as.Date("2022-01-01"))

ggplot(df, aes(x = reorder(Country,Bps), y = Bps, fill = Direction)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title = element_blank(),
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        plot.subtitle = element_text(face = "italic", size = 10, hjust = 0.32),
        axis.text = element_text(size = 15)) +
  scale_fill_manual(values = c("red","blue")) +
  labs(
    title = "The most recent change in interest rates in basis points",
    subtitle = "It includes the countries that decided on interest rates in 2022"
  )

```

Only 4 of the 73 central banks that made interest rate decisions in 2022 decided to cut rates, according to the list. China, Russia, Uzbekistan, and Turkey are among them.

Interest rates play an important role in determining the value of a currency. As a result, I'd like to demonstrate how the USDTRY reacts to interest rate decisions made by the CBRT. The data you can access by downloading the *post37.xlsx* file from [here](https://github.com/urazakgul/rblog/tree/main/data) is from Reuters.

The rates for the USDTRY and historical interest rates are shown separately below.

```{r fig.width=10, fig.height=5}

interest_rates <- readxl::read_excel("data.xlsx", sheet = "Policy Interest Rates") %>% 
  mutate(
    date = lubridate::ymd(paste0(word(date,3,3)," ",word(date,2,2)," ",word(date,1,1)))
  ) %>% 
  arrange(date)

ggplot(interest_rates, aes(x = date, y = policy_interest_rate)) +
  geom_line(size = 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 20),
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
  labs(
    title = "Interest Rate"
  )

```

```{r fig.width=10, fig.height=5}

usdtry_rates <- readxl::read_excel("data.xlsx", sheet = "USDTRY") %>% 
  mutate(date = as.Date(date)) %>% 
  arrange(date)

ggplot(usdtry_rates, aes(x = date, y = usdtry)) +
  geom_line(size = 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 20),
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
  labs(
    title = "USDTRY"
  )

```

If we normalize the data in the 0-1 range and combine the two variables, we get the following.

```{r fig.width=10, fig.height=5}

normalized <- function(x){
  
  (x - min(x)) / (max(x) - min(x))
  
}

df_join <- interest_rates %>% 
  left_join(usdtry_rates, by = "date") %>% 
  rename(
    "Interest Rate"=2,
    "USDTRY"=3
  ) %>% 
  mutate_at(
    vars(-date), function(x) normalized(x)
  ) %>% 
  pivot_longer(!date, names_to = "vars", values_to = "vals")

ggplot(df_join, aes(x = date, y = vals, group = vars, color = vars)) +
  geom_line(size = 1) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "top",
        legend.text = element_text(size = 20),
        legend.key.width = unit(3,"cm"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 20)) +
  scale_color_manual(values = c("red","blue"))

```

We can now look at how the CBRT's interest rate decisions affect the USDTRY.

```{r fig.width=15, fig.height=10}

master <- usdtry_rates %>% 
  left_join(interest_rates, by = "date")

changes <- data.frame(
  "DecisionDate" = interest_rates$date[-nrow(interest_rates)],
  "InterestRates" = interest_rates$policy_interest_rate[-nrow(interest_rates)],
  "FirstER" = NA,
  "SecondER" = NA
)

firstER <- NULL
secondER <- NULL

for(i in 1:nrow(master)){
  
  if(!(is.na(master$policy_interest_rate[i]))){
    
    firstER <- c(firstER,master$usdtry[i])
    secondER <- c(secondER,master$usdtry[(i-1)])
    
  }
  
  if(i == nrow(master)){
      firstER <- firstER[-length(firstER)]
      changes$FirstER <- firstER
      changes$SecondER <- secondER
      changes$Bps <- lag((lead(changes$InterestRates) - changes$InterestRates)*10000)
      changes$ER <- (changes$SecondER / changes$FirstER - 1) * 100

      changes <- changes %>%
        mutate(
          "Status" = case_when(
            Bps > 0 & ER > 0 ~ "IR Raises; ER Up",
            Bps > 0 & ER < 0 ~ "IR Raises; ER Down",
            Bps > 0 & ER == 0 ~ "IR Raises; ER Notr",
            Bps < 0 & ER < 0 ~ "IR Cuts; ER Down",
            Bps < 0 & ER > 0 ~ "IR Cuts; ER Up",
            Bps < 0 & ER == 0 ~ "IR Cuts; ER Notr",
            Bps == 0 & ER > 0 ~ "IR Keeps Constant; ER Up",
            Bps == 0 & ER < 0 ~ "IR Keeps Constant; ER Down",
            Bps == 0 & ER == 0 ~ "IR Keeps Constant; ER Notr"

          )
        ) %>%
        na.omit() %>%
        mutate(
          "Year" = format(DecisionDate,"%Y")
        )

    }
  
}

changes %>% 
  ggplot(aes(x = Bps, y = ER, color = Status)) +
  geom_point(size = 4, show.legend = FALSE) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  labs(
    x = "Basis Points",
    y = "Return"
  ) +
  theme(strip.text = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 15)) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~Year, scales = "free")

```

If we summarize the above graph;

```{r fig.width=10, fig.height=5}

changes %>% 
  group_by(Status) %>% 
  summarise(N = n()) %>% 
  ggplot(aes(x = reorder(Status, N), y = N, fill = N)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "orange", high = "red") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 15))

changes %>% 
  group_by(Status) %>% 
  summarise(N = n()) %>% 
  arrange(desc(N)) %>% 
  kableExtra::kbl(align = "c") %>% 
  kableExtra::kable_paper(full_width = F)

```

It seems that the CBRT will continue to cut interest rates in the coming months. As of 2017, the Turkish Lira depreciated until the next interest rate cut decision by the CBRT in 12 of the 14 cases.

If we visualize the movements of USDTRY with an initial value of 100 until the next CBRT interest rate decision, we get the following result.

```{r fig.width=13, fig.height=7, preview=TRUE}

master <- usdtry_rates %>% 
  left_join(interest_rates, by = "date") %>% 
  mutate(date = as.Date(date))

master$group <- NA

k <- 0

for(j in 1:nrow(master)){
  
  date <- master$date[j]
  
  if(date %in% interest_rates$date){
    
    master$group[j] <- paste0("G",k)
    k = k + 1
    
  }
  
  if(j == nrow(master)){
    
    master <- master %>% 
      mutate(group = zoo::na.locf(group)) %>% 
      filter(date >= as.Date("2017-03-16")) %>% 
      rename("DecisionDate"=1) %>% 
      left_join(changes[,c(1,7,8)], by = "DecisionDate") %>% 
      mutate(Status = word(Status,1,sep = "\\;"),
             Status = zoo::na.locf(Status),
             Year = zoo::na.locf(Year)) %>% 
      group_by(group) %>% 
      mutate(RN = row_number()) %>% 
      filter(DecisionDate != interest_rates$date[nrow(interest_rates)]) %>% 
      ungroup()
    
    master2 <- data.frame()
    
    for(m in 1:nrow(changes)){
      
      group_filtered <- master %>% 
        filter(group == paste0("G",m))
      
      minDateER <- group_filtered %>% 
        filter(DecisionDate == min(DecisionDate)) %>% 
        pull(usdtry)
      
      group_filtered$initialValue <- minDateER
      group_filtered$index <- group_filtered$usdtry / group_filtered$initialValue * 100
      
      master2 <- master2 %>% 
        bind_rows(group_filtered)
      
    }
    
  }
  
}

ggplot(master2, aes(x = RN, y = index, group = group, color = Status)) +
  geom_line(size = 1, show.legend = FALSE) +
  geom_hline(yintercept = 100, size = 1) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text = element_text(size = 15),
        plot.title = element_text(size = 15, face = "italic"),
        plot.caption = element_text(size = 10, face = "italic"),
        strip.text = element_text(size = 20)) +
  scale_color_manual(values = c("red","orange","blue")) +
  facet_wrap(~Status) +
  labs(
    title = "USDTRY exchange rate movement until the next CBRT's interest rate decision as of 2017",
    subtitle = "",
    x = "Number of Days",
    caption = "It should be noted that the CBRT's interest rate decision is not the only one that has an impact on the exchange rate."
  )

```

I'd like to remind you that, while variable interest rates can have an impact on exchange rates, domestic risks can also have an impact. Furthermore, the government is spending billions of dollars to suppress the USDTRY.

The codes that were used in the post are given below.

```{r echo=TRUE, eval=FALSE}

library(rvest)
library(tidyverse)

url <- "https://www.cbrates.com/"

df <- read_html(url) %>% 
  html_table() %>% 
  .[[4]] %>% 
  slice(-c(1:2)) %>% 
  select(-c(1,4)) %>% 
  mutate(
    "Date" = lubridate::mdy(gsub("[\\(\\)]", "", regmatches(X3, gregexpr("\\(.*?\\)", X3)))),
    "Country" = str_squish(word(X3,1,sep = "\\|")),
    "Interest Rate" = as.numeric(word(X2,1,1)),
    "Bps" = str_squish(gsub("[\\(\\)]", "", regmatches(X2, gregexpr("\\(.*?\\)", X2)))),
    "Direction" = word(Bps,1,1),
    "Bps" = as.numeric(word(str_squish(gsub("[\\(\\)]", "", regmatches(X2, gregexpr("\\(.*?\\)", X2)))),2,2)),
    "Bps" = case_when(
      Direction == "+" ~ Bps*100,
      Direction == "-" ~ Bps*(-100)
    )
  ) %>% 
  select(3:7) %>% 
  na.omit() %>% 
  filter(Date >= as.Date("2022-01-01"))

ggplot(df, aes(x = reorder(Country,Bps), y = Bps, fill = Direction)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title = element_blank(),
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        plot.subtitle = element_text(face = "italic", size = 10, hjust = 0.32),
        axis.text = element_text(size = 15)) +
  scale_fill_manual(values = c("red","blue")) +
  labs(
    title = "The most recent change in interest rates in basis points",
    subtitle = "It includes the countries that decided on interest rates in 2022"
  )

interest_rates <- readxl::read_excel("data.xlsx", sheet = "Policy Interest Rates") %>% 
  mutate(
    date = lubridate::ymd(paste0(word(date,3,3)," ",word(date,2,2)," ",word(date,1,1)))
  ) %>% 
  arrange(date)

ggplot(interest_rates, aes(x = date, y = policy_interest_rate)) +
  geom_line(size = 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 20),
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
  labs(
    title = "Interest Rate"
  )

usdtry_rates <- readxl::read_excel("data.xlsx", sheet = "USDTRY") %>% 
  mutate(date = as.Date(date)) %>% 
  arrange(date)

ggplot(usdtry_rates, aes(x = date, y = usdtry)) +
  geom_line(size = 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 20),
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) +
  labs(
    title = "USDTRY"
  )

normalized <- function(x){
  
  (x - min(x)) / (max(x) - min(x))
  
}

df_join <- interest_rates %>% 
  left_join(usdtry_rates, by = "date") %>% 
  rename(
    "Interest Rate"=2,
    "USDTRY"=3
  ) %>% 
  mutate_at(
    vars(-date), function(x) normalized(x)
  ) %>% 
  pivot_longer(!date, names_to = "vars", values_to = "vals")

ggplot(df_join, aes(x = date, y = vals, group = vars, color = vars)) +
  geom_line(size = 1) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "top",
        legend.text = element_text(size = 20),
        legend.key.width = unit(3,"cm"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 20)) +
  scale_color_manual(values = c("red","blue"))

master <- usdtry_rates %>% 
  left_join(interest_rates, by = "date")

changes <- data.frame(
  "DecisionDate" = interest_rates$date[-nrow(interest_rates)],
  "InterestRates" = interest_rates$policy_interest_rate[-nrow(interest_rates)],
  "FirstER" = NA,
  "SecondER" = NA
)

firstER <- NULL
secondER <- NULL

for(i in 1:nrow(master)){
  
  if(!(is.na(master$policy_interest_rate[i]))){
    
    firstER <- c(firstER,master$usdtry[i])
    secondER <- c(secondER,master$usdtry[(i-1)])
    
  }
  
  if(i == nrow(master)){
      firstER <- firstER[-length(firstER)]
      changes$FirstER <- firstER
      changes$SecondER <- secondER
      changes$Bps <- lag((lead(changes$InterestRates) - changes$InterestRates)*10000)
      changes$ER <- (changes$SecondER / changes$FirstER - 1) * 100

      changes <- changes %>%
        mutate(
          "Status" = case_when(
            Bps > 0 & ER > 0 ~ "IR Raises; ER Up",
            Bps > 0 & ER < 0 ~ "IR Raises; ER Down",
            Bps > 0 & ER == 0 ~ "IR Raises; ER Notr",
            Bps < 0 & ER < 0 ~ "IR Cuts; ER Down",
            Bps < 0 & ER > 0 ~ "IR Cuts; ER Up",
            Bps < 0 & ER == 0 ~ "IR Cuts; ER Notr",
            Bps == 0 & ER > 0 ~ "IR Keeps Constant; ER Up",
            Bps == 0 & ER < 0 ~ "IR Keeps Constant; ER Down",
            Bps == 0 & ER == 0 ~ "IR Keeps Constant; ER Notr"

          )
        ) %>%
        na.omit() %>%
        mutate(
          "Year" = format(DecisionDate,"%Y")
        )

    }
  
}

changes %>% 
  ggplot(aes(x = Bps, y = ER, color = Status)) +
  geom_point(size = 4, show.legend = FALSE) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  labs(
    x = "Basis Points",
    y = "Return"
  ) +
  theme(strip.text = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 15)) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~Year, scales = "free")

changes %>% 
  group_by(Status) %>% 
  summarise(N = n()) %>% 
  ggplot(aes(x = reorder(Status, N), y = N, fill = N)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "orange", high = "red") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 15))

master <- usdtry_rates %>% 
  left_join(interest_rates, by = "date") %>% 
  mutate(date = as.Date(date))

master$group <- NA

k <- 0

for(j in 1:nrow(master)){
  
  date <- master$date[j]
  
  if(date %in% interest_rates$date){
    
    master$group[j] <- paste0("G",k)
    k = k + 1
    
  }
  
  if(j == nrow(master)){
    
    master <- master %>% 
      mutate(group = zoo::na.locf(group)) %>% 
      filter(date >= as.Date("2017-03-16")) %>% 
      rename("DecisionDate"=1) %>% 
      left_join(changes[,c(1,7,8)], by = "DecisionDate") %>% 
      mutate(Status = word(Status,1,sep = "\\;"),
             Status = zoo::na.locf(Status),
             Year = zoo::na.locf(Year)) %>% 
      group_by(group) %>% 
      mutate(RN = row_number()) %>% 
      filter(DecisionDate != interest_rates$date[nrow(interest_rates)]) %>% 
      ungroup()
    
    master2 <- data.frame()
    
    for(m in 1:nrow(changes)){
      
      group_filtered <- master %>% 
        filter(group == paste0("G",m))
      
      minDateER <- group_filtered %>% 
        filter(DecisionDate == min(DecisionDate)) %>% 
        pull(usdtry)
      
      group_filtered$initialValue <- minDateER
      group_filtered$index <- group_filtered$usdtry / group_filtered$initialValue * 100
      
      master2 <- master2 %>% 
        bind_rows(group_filtered)
      
    }
    
  }
  
}

ggplot(master2, aes(x = RN, y = index, group = group, color = Status)) +
  geom_line(size = 1, show.legend = FALSE) +
  geom_hline(yintercept = 100, size = 1) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text = element_text(size = 15),
        plot.title = element_text(size = 15, face = "italic"),
        plot.caption = element_text(size = 10, face = "italic"),
        strip.text = element_text(size = 20)) +
  scale_color_manual(values = c("red","orange","blue")) +
  facet_wrap(~Status) +
  labs(
    title = "USDTRY exchange rate movement until the next CBRT's interest rate decision as of 2017",
    subtitle = "",
    x = "Number of Days",
    caption = "It should be noted that the CBRT's interest rate decision is not the only one that has an impact on the exchange rate."
  )

```