---
title: "Scraping Data from Wikipedia Tables"
description: |
  How to scrape and clean wikipedia tables.
author:
  - name: Uraz Akgül
date: 2022-09-18
output:
  distill::distill_article:
    self_contained: false
categories:
  - Web
  - Data Manipulation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I've scraped many websites, and scraping Wikipedia is by far the simplest. This post will teach you how to scrape tables from Wikipedia and how to clean the data that has been scraped.

The two urls we'll use are given below.

i. https://en.wikipedia.org/wiki/Democracy_Index

*The index is based on 60 indicators grouped in five different categories, measuring pluralism, civil liberties and political culture.*

ii. https://en.wikipedia.org/wiki/Corruption_Perceptions_Index

*The index which ranks countries by their perceived levels of public sector corruption, as determined by expert assessments and opinion surveys.*

For more information, please see the URLs above.

The index scores of countries participating in the Shanghai Cooperation Organization Summit in 2022 will be compared to those of countries that are not members of the organization.

```{r}

library(rvest)
library(tidyverse)

```

```{r}

url1 <- "https://en.wikipedia.org/wiki/Democracy_Index"
url2 <- "https://en.wikipedia.org/wiki/Corruption_Perceptions_Index"

participants <- c(
  "China",
  "Russia",
  "Kyrgyzstan",
  "Tajikistan",
  "Kazakhstan",
  "Uzbekistan",
  "India",
  "Pakistan",
  "Iran",
  "Turkmenistan",
  "Mongolia",
  "Belarus",
  "Turkey",
  "Azerbaijan"
)

```

While scraping the Democracy Index table, we followed the steps outlined below.

i. Read HTML code using the read_html() function

ii. Get tables using the html_table() function

iii. The 6th table in the list is the one we want (by running the first two lines of code, you can see all of the tables)

iv. Select country and score columns

v. Divide the countries based on the previously stated criteria

vi. Rename score column

```{r}

democracy <- read_html(url1) %>% 
  html_table() %>% 
  .[[6]] %>% 
  select(3,5) %>% 
  mutate(
    Shanghai = ifelse(
      Country %in% participants, "Shanghai", "Non-Shanghai"
    )
  ) %>% 
  rename(
    "Democracy"=2
  )

```

While scraping the Corruption Index table, we followed the steps outlined below.

i. Read HTML code using the read_html() function

ii. Get tables using the html_table() function

iii. The 5th table in the list is the one we want (by running the first two lines of code, you can see all of the tables)

iv. Select country and score columns

v. Remove the first line

vi. Rename country and score columns

vii. Convert character to numeric for the score column

```{r}

corruption <- read_html(url2) %>% 
  html_table() %>% 
  .[[5]] %>% 
  select(2,3) %>% 
  slice(-1) %>% 
  rename(
    "Country"=1,
    "Corruption"=2
  ) %>% 
  mutate(
    Corruption = as.numeric(Corruption)
  )

```

The inner_join() function can be used to join common countries from two tables.

```{r}

master <- democracy %>% 
  inner_join(corruption, by = "Country") %>% # the common rows between two tables
  select(Country,Democracy,Corruption,Shanghai)

```

As an example, the ten countries in the table are listed below.

```{r echo=FALSE}

master %>% 
  head(., 10) %>% 
  kableExtra::kbl(align = "c") %>% 
  kableExtra::kable_paper(full_width = F) %>% 
  kableExtra::column_spec(1, width = "10em", background = "yellow", bold = T) %>% 
  kableExtra::column_spec(2, width = "10em", background = "gray", color = "white", bold = T) %>% 
  kableExtra::column_spec(3, width = "10em", background = "gray", color = "white", bold = T) %>% 
  kableExtra::column_spec(4, width = "10em", background = "red", color = "white", bold = T)

```

Let's visualize the data in the last step.

```{r fig.width=15, fig.height=10, preview=TRUE}

ggplot(master, aes(x = Democracy, y = Corruption, color = Shanghai)) +
  geom_point(size = 5, alpha = .5) +
  ggrepel::geom_text_repel(data = master %>% filter(Shanghai == "Shanghai"),
                           aes(label = Country), size = 7, show.legend = FALSE) +
  geom_vline(xintercept = 5, linetype = "dashed") +
  geom_hline(yintercept = 50, linetype = "dashed") +
  ggthemes::theme_fivethirtyeight() +
  theme(
    axis.title = element_text(size = 25),
    axis.text = element_text(size = 20),
    legend.title = element_blank(),
    legend.position = "top",
    legend.text = element_text(size = 20)
  ) +
  scale_color_manual(values = c("gray60","red"))

```