---
title: "Close-to-Close vs High-Low (Parkinson) Volatility"
description: |
  Close-to-Close and High-Low (Parkinson) volatility types are compared.
author:
  - name: Uraz Akgül
date: 2022-09-27
output:
  distill::distill_article:
    self_contained: false
categories:
  - Finance
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Volatility can be measured in a variety of ways, and in this post I'd like to compare two of them.

The formulas for the two volatility methods mentioned in the title are shown below.

**Close-to-Close Volatility:**

Close-to-close historical volatility is calculated using only the closing prices of stocks.

$\sigma_{cl} = \sqrt{\frac{1}{T-1}\sum_{t=1}^{T}(r_t-\bar{r})^2}$

where:

$T:$ Number of days in the sample period

$r_t:$ Return on day t

$\bar{r}:$ Mean return

**High-Low or Parkinson Volatility:**

Parkinson volatility is a measure of volatility that takes into account the stock's high and low prices for the day.

$\sigma_{hl} = \sqrt{\frac{1}{4Tln2}\sum_{t=1}^{T}ln(\frac{H_t}{L_t})^2}$

where:

$T:$ Number of days in the sample period

$H_t:$ High price on day t

$L_t:$ Low price on day t

We'll use two methods to calculate the volatility of the stocks in the BIST30 index. The stock list is available on [PDP](https://www.kap.org.tr/en/Endeksler), which stands for Public Disclosure Platform. The data were obtained from Reuters, which you can access by downloading the *post34.xlsx* file from [here](https://github.com/urazakgul/rblog/tree/main/data).

```{r}

library(tidyverse)

```

```{r}

df <- readxl::read_excel("data.xlsx")

sampleSize <- nrow(df)
stockList <- unique(gsub("_Open|_High|_Low|_Close","",names(df)))[-1]

```

*Close-to-Close Volatility:*

```{r}

close_to_close <- data.frame()

for(i in 1:31){
  
  cl <- df %>% 
    select(contains(paste0(stockList[i],"_Close"))) %>% 
    rename("Close"=1) %>% 
    mutate(
      Return = lag(log(lead(Close) / Close)),
      Mean = mean(Return, na.rm = TRUE),
      ReturnMean2 = (Return - Mean)^2
    ) %>% 
    pull(ReturnMean2) %>% 
    sum(., na.rm = TRUE) / (sampleSize - 2) %>% # 31.12.2021 is not included & T - 1 = T - 2
    sqrt(.) * 100
  
  tbl_cl <- data.frame(
    "Ticker" = stockList[i],
    "ClosetoCloseVol" = cl
  )
  
  close_to_close <- close_to_close %>% 
    bind_rows(tbl_cl)
    
}

```

*High-Low or Parkinson Volatility:*

```{r}

high_low <- data.frame()

for(j in 1:31){
  
  hl <- df %>% 
    select(contains(c(paste0(stockList[j],"_High"),paste0(stockList[j],"_Low")))) %>% 
    rename("High"=1, "Low"=2) %>% 
    mutate(HL = log(High/Low)^2) %>% 
    pull(HL) %>% 
    sum(.) / (4 * (sampleSize - 1) * log(2)) %>% # 31.12.2021 is not included = T - 1
    sqrt(.) * 100
  
  tbl <- data.frame(
    "Ticker"= stockList[j],
    "HighLowVol" = hl
  )
  
  high_low <- high_low %>% 
    bind_rows(tbl)
  
}

```

```{r}

master <- close_to_close %>% 
  left_join(high_low, by = "Ticker")

```

```{r fig.width=10, fig.height=7, preview=TRUE}

ggplot(master, aes(x = ClosetoCloseVol, y = HighLowVol)) +
  geom_point(size = 5, alpha = .3) + 
  geom_point(data = master %>% filter(Ticker == "XU030"), color = "red", size = 5) +
  ggrepel::geom_text_repel(aes(label = Ticker), size = 4, nudge_x = 0.05) +
  hrbrthemes::theme_ipsum() +
  labs(
    title = "The Most and Least Volatile Stocks Among the BIST30 (XU030)",
    x = "Close-to-Close Volatility",
    y = "High and Low or Parkinson Volatility"
  )

```